---
title: "Malaria ML Model DashBoard(Kenya)"
author: "D.K.Muriithi"
date: "`r Sys.Date()`"
output: 
      html_document:  
  word_document: default
  pdf_document: default
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 4,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	comment = NA)
```

## Confirmation and setting of working directory
```{r}
setwd("~/TMMS2024")
```

## Installation and loading of necessary packages/libraries

## Loading libraries
```{r}
library(caret) #for training machine learning models
library(psych) ##for description of  data
library(ggplot2) ##for data visualization
library(caretEnsemble)##enables the creation of ensemble models
library(tidyverse) ##for data manipulation
library(mlbench)  ## For benchmarking ML Models
library(flextable) ## To create and style tables
library(mltools) #for hyperparameter tuning
library(tictoc) #for determining the time taken for a model to run
library(ROSE)  ## for random oversampling
library(smotefamily) ## for smote sampling
library(ROCR) ##For ROC curve
library(pROC) ## For visualizing, smoothing, and comparing ROC curves
library(e1071) ## For statistical modeling and  machine learning tasks
library(class) ## For classification using k-Nearest Neighbors and other methods
library(caTools) ## For splitting data into training and testing sets
library(MASS) ## Provides plotting functions and datasets
library(ISLR) ## for practical applications of statistical learning methods
library(boot) ## Useful for performing bootstrap resampling
library(cvTools) ## Contains functions for cross-validation, bootstrapping, and other resampling methods
```

## Loading the given Malaria data

https://statistics.knbs.or.ke/nada/index.php/catalog/111/related-materials

```{r}
mdata = read.csv("final_malaria_survey_data.csv", header = TRUE)
head(mdata)
```
## Exporatory of the dataset
```{r}
#dim(mdata)      ## View the Dimension of the Data
#names(mdata)     ## View the variable/features/column names
#summary(mdata)    ## Descriptive Statistics
#describe(mdata)   ## Descriptive Statistics
#sum(is.na(mdata))  ## Check for missing data
#na.omit(mdata)     ## Remove rows with any missing values
#is.na(mdata$Malaria.Test)  ## checks for missing values in the Malaria.Test column of your data frame
```

## Note: For the purpose of this training: It is assumed that the data is already clean and preprocessed 

# Factor the Target variable
```{r}
mdata$Malaria.Test <- as.factor(mdata$Malaria.Test)
```

# Plot Target variable using ggplot function
```{r}
ggplot(mdata, aes(x = Malaria.Test)) + 
  geom_bar(fill= c("green", "red")) + 
  labs(x = "Malaria Test", 
       y = "Respondent",
       tittle = "Malaria Diagnosis Results",
       caption = "Source: KNBS 2021 Data") +
    theme_classic()
```

# Check for zero variance predictors:
```{r}
nzv <- nearZeroVar(mdata[,-14], saveMetrics = TRUE) ## Function called nearZeroVar and captures its output in the variable nzv
```

# Remove nzv
```{r}
mdata1 <- mdata[, !nzv$nzv] ## Removing features with little to no variation
dim(mdata1)
```

## DATA PARTITION FOR MACHINE LEARNING
```{r}
set.seed(123)
index = sample(2, nrow(mdata1),replace =T, prob=c(0.70,0.30))
train = mdata1[index ==1,]
test = mdata1[index ==2,]
```

# Get the dimensions of your train and test data
```{r}
dim(train)
dim(test)
```

# Prepare training scheme for cross-validation 

# Cross-validation
This involves splitting the data into multiple subsets (folds), training the model on some folds, and testing it on the remaining fold. The process is repeated for each fold
# Repeated cross-validation 
This involves performing cross-validation multiple times to reduce the variability of the results reducing the likelihood of overfitting

```{r}
control <- trainControl(method = "repeatedcv", number = 10, repeats=5, sampling ='smote')# SMOTE sampling
```

# Train a Random Forest model
# Random Forests 
This is an ensemble learning method that combines multiple decision trees to improve prediction accuracy and reduce variance.

# mtry
This parameter controls the number of features randomly chosen as candidates for splitting a node in each tree.
```{r}
tuneGrid_rf <- expand.grid(mtry = c(2, 4, 6, 8, 12))
```


```{r}
set.seed(123)
tic()
RFModel <- train(factor(Malaria.Test)~., 
                 data=train, 
                 method="rf", 
                 trControl=control, 
                 tuneGrid=tuneGrid_rf,
                 na.action = na.omit)
toc()
#RFModel
#RFModel$results
```

```{r}
plot(RFModel)
```


```{r}
# Prediction on test data set using RF model
RFpred=predict(RFModel,newdata = test)

# Evaluate RF model performance metrics
RF_CM<- confusionMatrix(RFpred, as.factor(test$Malaria.Test), positive = "Positive", mode='everything')
M2<- RF_CM$byClass[c(1, 2, 5, 7, 11)]
M2

# Ploting Random Forest confusion matrix
fourfoldplot(RF_CM$table, col=rainbow(4), main="RF Confusion Matrix") #RF Confusion Matrix 4fold plot

# Show relative importance of features
# Plot using R base function
plot(varImp(RFModel, scale=T))

# Alternatively
vip::vip(RFModel)

# Alternatively using ggplot function
var_imp <-varImp(RFModel)
ggplot(var_imp, aes(x = reorder(Variable, Importance), y = importance)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for RF Model")
```

# Create ROC curve for RF model
```{r}
# Make predictions on the test set using type='prob'
predrf <- predict(RFModel, newdata = test, type = "prob")
# Create a prediction object needed by ROCR
pred_rf <- prediction(predrf[, "Positive"], test$Malaria.Test)
# Calculate performance measures like ROC curve
perf_rf <- performance(pred_rf, "tpr", "fpr")
# Plot the ROC curve
plot(perf_rf, colorize = TRUE, main = "ROC Curve-Random Forest")
# Compute AUC
auc_value <- performance(pred_rf, "auc")@y.values[[1]]
auc_label <- paste("AUC =", round(auc_value, 2))
# Add AUC value as text on the plot
text(0.5, 0.3, auc_label, col = "red", cex = 1.5)  # Adjust position
```


```{r}
# Save the model to a file
saveRDS(RFModel, "RANDOM_FOREST.rds")
#View(mdata1)
```


```{r}
library(shiny)

# Load the trained model
RFModel <- readRDS("RANDOM_FOREST.rds")

# Define UI
ui <- fluidPage(
  titlePanel("ML Model Dashboard"),
  sidebarLayout(
    sidebarPanel(
      numericInput("Region","Region",value = 1) ,
      numericInput("County","County" , value =  19) , 
      numericInput("Zones", "Zones", value = 1 ) ,   
      numericInput("Has.mosquito.bed.net.for.sleeping" ,"Has.mosquito.bed.net.for.sleeping",value = 1 ),
      numericInput("Children.under.5.who.slept.under.mosquito.net.last.night","Children.under.5.who.slept.under.mosquito.net.last.night", value = 1),
      numericInput("Number.of.Mosquito.bed.nets","Number.of.Mosquito.bed.nets",value = 1),
      numericInput("Number.of.children.under.mosquito.bed.nets.last.night", "Number.of.children.under.mosquito.bed.nets.last.night", value =1),
      numericInput("Anaemia.Level", "Anaemia.Level",value = 1 ) , 
      numericInput("Mothers.highest.educational.level" , "Mothers.highest.educational.level" , value = 1), 
      numericInput("The.type.of.mosquito.net.the.person.slept.under.last.night","The.type.of.mosquito.net.the.person.slept.under.last.night",value = 1),
      numericInput("Presence.of.species..Falciparum","Presence.of.species..Falciparum",value =  1 ) ,
      numericInput("Presence.of.species..Malariae","Presence.of.species..Malariae",value = 0 ),
      numericInput("Presence.of.species..Ovale" , "Presence.of.species..Ovale" ,value =  0 ) ,  
      actionButton("predict", "Predict")
    ),
    mainPanel(
      h3("Prediction:"),
      verbatimTextOutput("result")
    )
  )
)

# Define server logic
server <- function(input, output) {
  observeEvent(input$predict, {
    # Create a new data frame with user input
    new_data <- data.frame(
      Region = input$Region,
      County = input$County,
      Zones = input$Zones,
      Has.mosquito.bed.net.for.sleeping= input$Has.mosquito.bed.net.for.sleeping,
      Children.under.5.who.slept.under.mosquito.net.last.night= input$Children.under.5.who.slept.under.mosquito.net.last.night,
      Number.of.Mosquito.bed.nets = input$Number.of.Mosquito.bed.nets,
      Number.of.children.under.mosquito.bed.nets.last.night = input$Number.of.children.under.mosquito.bed.nets.last.night,
      Anaemia.Level = input$Anaemia.Level,
      Mothers.highest.educational.level = input$Mothers.highest.educational.level,
      The.type.of.mosquito.net.the.person.slept.under.last.night = input$The.type.of.mosquito.net.the.person.slept.under.last.night,
      Presence.of.species..Falciparum = input$Presence.of.species..Falciparum,
      Presence.of.species..Malariae = input$Presence.of.species..Malariae,
      Presence.of.species..Ovale= input$Presence.of.species..Ovale
  )
    
    # Predict the species
    prediction <- predict(RFModel, new_data)
    
    # Show the result
    output$result <- renderText({ as.character(prediction) })
  })
}
```


```{r}

# Run the Shiny app
shinyApp(ui = ui, server = server)
```



                         
